---
title: "Model Selection - Grid Search"
output: html_document
---

Grid Search optimizes model hyperparameters - the parameters we choose.

The caret package is used because not only does it provide just about any ML algo you may want, it also simplifies hyperparameter optimization.

> Remember the caret package is great but using other packages sometimes provides options not available in caret for the same ML algorithm.

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(!require(easypackages)){install.packages("easypackages")}
library(easypackages)
packages("caTools", "e1071", "caret", "ElemStatLearn",  prompt = FALSE)
```

```{r}
# Importing the dataset
dataset = read.csv('./data/Social_Network_Ads.csv')
dataset = dataset[3:5]

# Encoding the target feature as factor
dataset$Purchased = factor(dataset$Purchased, levels = c(0, 1))

# Splitting the dataset into the Training set and Test set - caTools
set.seed(123)
split = sample.split(dataset$Purchased, SplitRatio = 0.75)
training_set = subset(dataset, split == TRUE)
test_set = subset(dataset, split == FALSE)

# Feature Scaling
training_set[-3] = scale(training_set[-3])
test_set[-3] = scale(test_set[-3])
```
```{r}
# Fitting Kernel SVM to the Training set - e1071
classifier = svm(formula = Purchased ~ ., data = training_set,
        type = 'C-classification', kernel = 'radial')

# Predicting the Test set results
y_pred = predict(classifier, newdata = test_set[-3])

# Making the Confusion Matrix
cm = table(test_set[, 3], y_pred)

# Applying k-Fold Cross Validation - caret
folds = createFolds(training_set$Purchased, k = 10)
cv = lapply(folds, function(x) {
  training_fold = training_set[-x, ]
  test_fold = training_set[x, ]
  classifier = svm(formula = Purchased ~ ., data = training_fold,
                   type = 'C-classification', kernel = 'radial')
  y_pred = predict(classifier, newdata = test_fold[-3])
  cm = table(test_fold[, 3], y_pred)
  accuracy = (cm[1,1] + cm[2,2]) / (cm[1,1] + cm[2,2] + cm[1,2] + cm[2,1])
  return(accuracy)
})
accuracy = mean(as.numeric(cv))
```

Now lets use the caret package to optimize the hyperparameters.
```{r message=FALSE}
# Applying Grid Search to find the best parameters - caret
# Automatically boostraps - smae as K=Fo;d so whhy not always Grid Search - RESEARCH THIS
classifier = train(form = Purchased ~ ., data = training_set, method = 'svmRadial')
classifier
classifier$bestTune # gives the best values directly
```

You could take the values from the Grid results and plug them directly into the first classifier we did previously to get the same result.
