---
title: "Visualizing Decision Tree Boundaries"
output:
  rmdformats::material:
    highlight: pygments
    code_folding: hide
---

`parttree` includes a set of simple functions for visualizing decision tree partitions in R with `ggplot2.` 

```{r}
# install.packages("remotes")
remotes::install_github("grantmcdermott/parttree")
library("parttree")
library("tidyverse")
```

Using the familiar `ggplot2` syntax, we can simply add decision tree boundaries to a plot of our data.

Train a decision tree on the famous Titanic data using `parsnip`. Then visualize the resulting partition / decision boundaries using the simple function `geom_parttree()`,=.

```{r message=FALSE, warning=FALSE}
library(parsnip)
library(titanic) ## Just for a different data set
set.seed(123) ## For consistent jitter

titanic_train$Survived = as.factor(titanic_train$Survived)

## Build our tree using parsnip (but with rpart as the model engine)
ti_tree =
  decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification") %>%
  fit(Survived ~ Pclass + Age, data = titanic_train)

## Plot the data and model partitions
titanic_train %>%
  ggplot(aes(x=Pclass, y=Age)) +
  geom_jitter(aes(col=Survived), alpha=0.7) +
  geom_parttree(data = ti_tree, aes(fill=Survived), alpha = 0.1) +
  theme_minimal()
```
Super awesome!

This visualization precisely shows where the trained decision tree thinks it should predict that the passengers of the Titanic would have survived (blue regions) or not (red), based on their age and passenger class (Pclass).