---
title: "Nested Logistic Examples"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The document uses `mlogit` to provide exampleds for nested logistic regression.  'mlogit` enables the estimation of the multinomial logit models with individual and/or alternative speci???c variables. The main extensions of the basic multinomial model (heteroscedastic, nested and random parameter models) are implemented.

Data sets used for multinomial logit estimation deals with some individuals that make one or a sequential choice of one alternative among a set of several alternatives. The determinants of these choices are variables that can be alternative speci???c or purely individual speci???c. Such data have therefore a speci???c structure that can be characterised by three indexes :

- the alternative - these do not work for auto - the values are the same
- the choice situation
- the individual (only relevant if we have repeated observations for the same individual)

Because `mlogit` requires the format of the data to be entered into the formula.  The information about the structure of the data can be explicitly indicated or, in part, guessed by `mlogit.data`. 

Data sets can have two di???erent shapes:

- wide shape: There is one row for each choice situation
- long shape : There is one row for each alternative and therefor, as many rows as there are alternatives for each choice situation

# Review Data

```{r message=FALSE, warning=FALSE}
if(!require(easypackages)){install.packages("easypackages")}
library(easypackages)
packages("dplyr", "mlogit", "AER", "readr",  prompt = FALSE)
```

## Data - Auto Valuation Sample
```{r}
autoVal <- read_csv("S:/AutoNestLog/sample1.csv", n_max = 5000)
fishing <- data("Fishing", package = "mlogit")
travelMode <- data("TravelMode", package="AER") 
```

## Data - Fishing
```{r}
glimpse(Fishing)
```

There are four ???shing modes (beach, pier, boat, charter), two alternative speci???c variables (price and catch) and one choice/individual speci???c variable (income)2. This "wide" format is suitable to store individual speci???c variables. 

## Data - Train
```{r}
glimpse(TravelMode)
```

There are four transport modes (air, train, bus and car) and most of the variable are alternative speci???c (wait, vcost, travel, gcost). The only individual speci???c variables are income and size. The advantage of this shape is that there are much fewer columns than in the wide format, the caveat being that values of income and size are repeated four times. `mlogit` deals with both format. It provides a `mlogit.data` function that take as ???rst argument a data.frame and returns a data.frame in "long" format with some information about the structure of the data. 

Note this is long format data.

## Data - autoVal

```{r}
glimpse(autoVal)
```

> The `autoVal` data appears to align more closely with the `TravelMode` data.

To ensure `autoVal` is correctly used by `mlogit`, transform the data to more closely align with the `TravelMode` data.

```{r}
autoVal <- autoVal %>% rename(mode = event_type, individual = person_id, choice = decision)

autoVal <- autoVal %>% select(-(active:skip))

glimpse(autoVal)
```

To further convert the `autoVal` data to look like the `TravelMod` data, change a few features to factors.

```{r}
autoVal <- autoVal %>% mutate_at(vars("individual", "choice", "mode"), funs(factor(.)))
glimpse(autoVal)
```

There are 3 event types (active (1), repo (2), skip (3)) and most of the variables are individual specific (fCB, aPTI, aDTI).  `LTV24` is an alternate specific variable.

# Modeling with `mlogit`

```{r}
#Test with and without varying parameter
myLogit_DF <-  mlogit.data(autoVal, choice = "choice", shape = "long", varying = "LTV24", 
                           chid.var = "individual", alt.var = "mode")
head(myLogit_DF)
```

```{r nestedLogit}
#MOB related value to choice?
myNest <- mlogit(choice ~ fCB + aPTI + aDTI + LTV24, autoVal, nests = list(Good = "1", Bad = c("2", "3")), unscaled = TRUE)
```


```{r scratchWork}
data("TravelMode",package="AER")
class(TravelMode)
str(TravelMode)

#individual: Factor indicating individual with levels 1 to 200.
#mode: Factor indicating travel mode with levels "car", "air", "train", or "bus".
#choice: Factor indicating choice with levels "no" and "yes".
#wait: Terminal waiting time, 0 for car.
#vcost: Vehicle cost component.
#travel: Travel time in the vehicle.
#gcost: Generalized cost measure.
#income: Household income.
#size: Party size.

TravelMode <- mlogit.data(TravelMode,choice="choice",shape="long", alt.var="mode",chid.var="individual")
str(TravelMode)

TravelMode$avinc <- with(TravelMode,(mode=='air')*income)

nl.TM <- mlogit(choice ~ wait + gcost + avinc, TravelMode, reflevel = "car", nests = list(fly = "air", ground = c("train", "bus", "car")), unscaled=TRUE)

summary(nl.TM)


#autoVal <- read_csv("S:/AutoNestLog/sample1.csv", n_max = 1440)
autoVal <- read_csv("S:/AutoNestLog/sample1.csv")
str(autoVal)#want individual, mode and choice to be factors just like TravelMode example
autoVal <- autoVal %>% rename(mode = event_type, individual = person_id, choice = decision)

#autoVal <- autoVal %>% select(-(active:skip), -LTV24)
autoVal <- autoVal %>% select(-(active:skip))

autoVal$mode[autoVal$mode == 1] <- "Active"
autoVal$mode[autoVal$mode == 2] <- "Repo"
autoVal$mode[autoVal$mode == 3] <- "Skip"

autoVal$choice[autoVal$choice == 1] <- "TRUE"
autoVal$choice[autoVal$choice == 0] <- "FALSE"

autoVal$fCB <- as.integer(autoVal$fCB)

autoVal <- autoVal %>% mutate_at(vars("individual", "choice", "mode"), funs(factor(.)))
autoVal <- select(autoVal, individual, mode, choice, everything())
autoVal <- as.data.frame(autoVal)

# class(autoVal)
# str(autoVal)

autoVal_logit <- mlogit.data(autoVal, choice = "choice", shape = "long", alt.var = "mode", chid.var = "individual")
str(autoVal_logit)

myNest <- mlogit(choice ~ MOB, autoVal_logit, reflevel = "Active", nests = list(Good = "Active", Bad = c("Repo", "Skip")), unscaled = TRUE)
# Error:  https://stackoverflow.com/questions/47012554/error-in-mlogit-package-system-is-computationally-singular-reciprocal-conditio

# Add skip feature per Anil 8/14/18 ############################################
autoVal <- read_csv("S:/AutoNestLog/sample1.csv")
autoVal <- autoVal %>% rename(mode = event_type, individual = person_id, choice = decision)
#autoVal <- autoVal %>% select(-(active:skip), -LTV24)
autoVal <- autoVal %>% select(-(active:repo))

autoVal$mode[autoVal$mode == 1] <- "Active"
autoVal$mode[autoVal$mode == 2] <- "Repo"
autoVal$mode[autoVal$mode == 3] <- "Skip"

autoVal$choice[autoVal$choice == 1] <- "TRUE"
autoVal$choice[autoVal$choice == 0] <- "FALSE"

autoVal <- autoVal %>% mutate_at(vars("fCB"), funs(as.integer(.)))
autoVal <- autoVal %>% mutate_at(vars("individual", "mode"), funs(factor(.)))
autoVal$choice <- as.logical(autoVal$choice)
autoVal <- select(autoVal, individual, mode, choice, everything())
autoVal <- as.data.frame(autoVal)

tmpDF <- autoVal %>% filter(choice == "TRUE") %>% group_by(mode) %>% summarise(count = n()) %>% 
                                      mutate(prop = count / sum(count))
# 88.2%active; 10.5% Repo; 1.2% Skip

# class(autoVal)
# str(autoVal)

autoVal_logit <- mlogit.data(autoVal, choice = "choice", shape = "long", alt.var = "mode", chid.var = "individual")
str(autoVal_logit)

myNest <- mlogit(choice ~ MOB + fCB + LTV24 + aPTI + aDTI + skip, autoVal_logit, reflevel = "Active", nests = list(Good = "Active", Bad = c("Repo", "Skip")), unscaled = TRUE)

myNest <- mlogit(choice ~ skip, autoVal_logit, reflevel = "Active", nests = list(Good = "Active", Bad = c("Repo", "Skip")), unscaled = TRUE)


#####################
mlogit(choice ~ fCB + aPTI + aDTI, data = autoVal_logit, reflevel = "Active")

model_tmp <- glm(choice ~ fCB + aPTI + aDTI, data = autoVal, family = "binomial")
```

```{r mLogit_explore}
## Cameron and Trivedi's Microeconometrics p.493 There are two
## alternative specific variables : price and catch one individual
## specific variable (income) and four fishing mode : beach, pier, boat,
## charter

data("Fishing", package = "mlogit")
glimpse(Fishing)

# There are four fishing modes (beach, pier, boat, charter), two alternative specific variables
# (price and catch) and one choice/individual specific variable (income). This wid"
# format is suitable to store individual specificc variables.

Fish <- mlogit.data(Fishing, varying = c(2:9), shape = "wide", choice = "mode")
glimpse(Fish)
## a pure "conditional" model

summary(mlogit(mode ~ price + catch, data = Fish))

## a pure "multinomial model"

summary(mlogit(mode ~ 0 | income, data = Fish))

## which can also be estimated using multinom (package nnet)

library("nnet")
summary(multinom(mode ~ income, data = Fishing))

## a "mixed" model

m <- mlogit(mode ~ price+ catch | income, data = Fish)
summary(m)

## same model with charter as the reference level

m <- mlogit(mode ~ price+ catch | income, data = Fish, reflevel = "charter")
summary(m)
## same model with a subset of alternatives : charter, pier, beach

m <- mlogit(mode ~ price+ catch | income, data = Fish,
            alt.subset = c("charter", "pier", "beach"))

## model on unbalanced data i.e. for some observations, some
## alternatives are missing

# a data.frame in wide format with two missing prices
Fishing2 <- Fishing
Fishing2[1, "price.pier"] <- Fishing2[3, "price.beach"] <- NA
mlogit(mode~price+catch|income, Fishing2, shape="wide", choice="mode", varying = 2:9)

# a data.frame in long format with three missing lines
data("TravelMode", package = "AER")
Tr2 <- TravelMode[-c(2, 7, 9),]
mlogit(choice~wait+gcost|income+size, Tr2, shape = "long",
       chid.var = "individual", alt.var="mode", choice = "choice")

## An heteroscedastic logit model

data("TravelMode", package = "AER")
hl <- mlogit(choice ~ wait + travel + vcost, TravelMode,
             shape = "long", chid.var = "individual", alt.var = "mode",
             method = "bfgs", heterosc = TRUE, tol = 10)

## A nested logit model

TravelMode$avincome <- with(TravelMode, income * (mode == "air"))
TravelMode$time <- with(TravelMode, travel + wait)/60
TravelMode$timeair <- with(TravelMode, time * I(mode == "air"))
TravelMode$income <- with(TravelMode, income / 10)

# Hensher and Greene (2002), table 1 p.8-9 model 5
TravelMode$incomeother <- with(TravelMode, ifelse(mode %in% c('air', 'car'), income, 0))
glimpse(TravelMode)
nl <- mlogit(choice~gcost+wait+incomeother, TravelMode,
             shape='long', alt.var='mode',
             nests=list(public=c('train', 'bus'), other=c('car','air')))

# same with a comon nest elasticity (model 1)
nl2 <- update(nl, un.nest.el = TRUE)

## a probit model
## Not run: 
pr <- mlogit(choice ~ wait + travel + vcost, TravelMode,
             shape = "long", chid.var = "individual", alt.var = "mode",
             probit = TRUE)

## End(Not run)


## a mixed logit model
## Not run: 
rpl <- mlogit(mode ~ price+ catch | income, Fishing, varying = 2:9,
              shape = 'wide', rpar = c(price= 'n', catch = 'n'),
              correlation = TRUE, halton = NA,
              R = 10, tol = 10, print.level = 0)
summary(rpl)
rpar(rpl)
cor.mlogit(rpl)
cov.mlogit(rpl)
rpar(rpl, "catch")
summary(rpar(rpl, "catch"))

## End(Not run)

# a ranked ordered model
data("Game", package = "mlogit")
g <- mlogit(ch~own|hours, Game, choice='ch', varying = 1:12,
            ranked=TRUE, shape="wide", reflevel="PC")


```




# Primary References

1. https://cran.r-project.org/web/packages/mlogit/vignettes/mlogit.pdf
2. Predicting with mlogit (harder than it should be: https://stats.stackexchange.com/questions/6702/predict-after-running-the-mlogit-function-in-r/60778

