---
title: 'RtextSummary Introduction'
output:
    rmdformats::readthedown:
      highlight: pygments
      code_folding: show
---

https://cran.r-project.org/web/packages/RtextSummary/RtextSummary.pdf

Build a text summary by extracting relevant sentences from your text. The training dataset should
consist of several documents, each document should have sentences separated by a period. While
fitting the model, the ’term frequency - inverse document frequency’ (TF-IDF) matrix that reflects
how important a word is to a document is calculated first. Then vector representations for words are
obtained from the ’global vectors for word representation’ algorithm (GloVe). While applying the
model on new data, the GloVe word vectors for each word are weighted by their TF-IDF weights
and averaged to give a sentence vector or a document vector. The magnitude of this sentence vector
gives the importance of that sentence within the document. Another way to obtain the importance of
the sentence is to calculate cosine similarity between the sentence vector and the document vector.
The output can either be at the sentence level (sentences and weights are returned) or at a document
level (the summary for each document is returned). It is useful to first get a sentence level output and
get quantiles of the sentence weights to determine a cutoff threshold for the weights. This threshold
can then be used in the document level output

```{r}
library(RtextSummary)
library(tidyverse)
```

```{r}
data("opinosis")

# the data is reduced to pass CRAN checks of <5 sec run-time
# delete the line below to build the model on the entire dataset
opinosis = opinosis[1:2,]%>%mutate(text = substr(text, 0, 10) )

```

```{r}
# 'stopwords_longlist' is a very long list of stopwords.
# it is not used in this example but can be useful for other datasets
data("stopwords_longlist")
```

```{r}
opinosis$text = stringr::str_replace_all(stringr::str_to_lower(opinosis$text),'[^a-z. ]','' )
```

```{r}
# -- the model will be fit at the sentence level, which works well for this dataset
# for other datasets, also try fitting at the document level by commenting out the two lines below
tempdf = opinosis%>% tidyr::separate_rows(text, sep = '\\.')
# ----------------------------------------
summary.model = TextSummary$new( stopword_list = c() )
summary.model$fit(tempdf$text)
# the parameters below work well for this dataset.
# For other datasets, try changing weight_method and avg_weight_by_word_count
df_sentence_level = summary.model$transform(opinosis, doc_id = 'topics', txt_col = 'text',
                                            summary_col = 'summary', weight_method = 'Magnitude',
                                            return_sentences = TRUE, avg_weight_by_word_count = TRUE)
# explore weight thresholds
quantile(df_sentence_level$wt, seq(0,1,0.1))

df_summary = summary.model$transform(opinosis, doc_id = 'topics', txt_col = 'text', summary_col = 'summary',
                                     weight_method = 'Magnitude', topN = 1, weight_threshold=quantile(df_sentence_level$wt, 0.3 ),
                                     return_sentences = FALSE, replace_char = '', avg_weight_by_word_count = TRUE)

```


