---
title: "ruintehol Example"
author: "Cliff W."
date: "4/23/2019"
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
    fig_caption: false
---

```{r}
if(!require(easypackages)){install.packages("easypackages")}
library(easypackages)
packages("tidyverse", "ruimtehol", prompt = T)
```

# Introduction

This R package is a wrapper around Starspace which provides a neural embedding model for doing the following on text:

- Text classification
- Learning word, sentence or document level embeddings
- Finding sentence or document similarity
- Ranking web documents
- Content-based recommendation (e.g. recommend text/music based on the content)
- Collaborative filtering based recommendation (e.g. recommend text/music based on interest)
- Identification of entity relationships

The list of R packages regarding text mining with R provided by BNOSAC has been steadily growing. This is the list of R packages maintained by BNOSAC.

- __udpipe__: tokenisation, lemmatisation, parts of speech tagging, dependency parsing, morphological feature extraction, sentiment scoring, keyword extraction, NLP flows
- __crfsuite__: named entity recognition, text classification, chunking, sequence modelling
- __textrank__: text summarisation
- __ruimtehol__: text classification, word/sentence/document embeddings, document/label similarities, ranking documengs, content based recommendation, collaborative filtering-based recommendation

# Example

Below is an example how the package can be used for multi-label classification on questions asked in Belgian parliament. Each question in parliament was labelled with several of one of the 1785 categories.

```{r}
data(dekamer, package = "ruimtehol")

glimpse(dekamer)
class(dekamer)
```

Each question in parliament was labelled with more than 1 category. There are 1785 categories in this dataset

Split up `question_theme.`  Create new new variable `question_themes` - plural.

```{r}
dekamer$question_themes <- strsplit(dekamer$question_theme, " +\\| +")# Finds spaced between wrds in a sentence
glimpse(dekamer)
head(dekamer$question_themes, 5)
```

Plain text of the question in parliament

```{r}
head(dekamer$question, 2)
dekamer$text <- strsplit(dekamer$question, "\\W") # Do a global search for non-word characters in a string
head(dekamer$question, 2)
```

```{r}
dekamer$text <- sapply(dekamer$text, FUN=function(x) paste(x, collapse = " "))
dekamer$text <- tolower(dekamer$text)
head(dekamer$text, 2)
```

# Build starspace model
```{r}

model <- embed_tagspace(x = dekamer$text, 
                        y = dekamer$question_themes, 
                        dim = 50, 
                        ngram = 3, loss = "hinge", similarity = "cosine", adagrad = TRUE,
                        early_stopping = 0.8, minCount = 2, 
                        thread = 4)
plot(model)
```

Get embeddings of the dictionary of words as well as the categories
```{r}
embedding_words  <- as.matrix(model, type = "words")
embedding_labels <- as.matrix(model, type = "label")
```

```{r}
## Find closest labels / predict
embedding_combination <- starspace_embedding(model, "federale politie patrouille", type = "document")
embedding_similarity(embedding_combination, embedding_labels, top_n = 3)
```

```{r}
predict(model, "de migranten komen naar europa, in asielcentra ...")

#$input
"de migranten komen naar europa, in asielcentra ..."
#$prediction
```

```{r}
dekamer$x <- strsplit(dekamer$question, "\\W")
glimpse(dekamer)
```
```{r}
dekamer$x <- sapply(dekamer$x, FUN = function(x) paste(setdiff(x, ""), collapse = " "))
glimpse(dekamer)
```
```{r}
dekamer$x <- tolower(dekamer$x)
dekamer$y <- strsplit(dekamer$question_theme, split = ",")
head(dekamer$y, 5)
```
```{r}
dekamer$y <- lapply(dekamer$y, FUN=function(x) gsub(" ", "-", x))
head(dekamer$y, 5)
```
```{r}
set.seed(123456789)
model <- embed_tagspace(x = dekamer$x, y = dekamer$y,
                        dim = 50, 
                        lr = 0.01, epoch = 40, loss = "softmax", adagrad = TRUE, 
                        similarity = "cosine", negSearchLimit = 50,
                        ngrams = 2, minCount = 2)
plot(model) 
```

```{r}
text <- c("de nmbs heeft het treinaanbod uitgebreid via onteigening ...",
          "de migranten komen naar europa de asielcentra ...")                   
predict(model, text, k = 3)  
predict(model, "koning filip", k = 10, type = "knn")
predict(model, "koning filip", k = 10, type = "embedding")
```

Reference

https://www.r-bloggers.com/starspace-for-nlp-nlproc/
