---
title: "textplot - Visualization of Text by bnosac"
output:
  rmdformats::readthedown:
    highlight: pygments
    code_folding: hide
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(!require(easypackages)){install.packages("easypackages")}
library(easypackages)
packages("udpipe", "textplot", "ggraph", "BTM", "concaveman", "Rgrapghviz", "qgraph", "glasso", prompt = FALSE)

# BTM - Biterm topic models for short text
# concaveman - Fast 2D Concave Hull Algorithm
## Algorithms that construct convex hulls of various objects have a broad range of applications in mathematics and computer science.
## In computational geometry, numerous algorithms are proposed for computing the convex hull of a finite set of points, 
## with various computational complexities.

setwd("~/Documents/GitHub/MachineLearning/Models/NLP_TextMining")
```

```{r installGraph, eval=FALSE}
# http://www.bioconductor.org/packages/release/bioc/html/graph.html
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

BiocManager::install("graph")
BiocManager::install("Rgraphviz")
```

```{r message=FALSE, warning=FALSE}
library(graph)
library(Rgraphviz)
```

# Introduction

The `texplot` R package allows one to visualise complex relations in texts. This is done by providing functionalities for displaying text co-occurrence networks, text correlation networks, dependency relationships as well as text clustering.

`textplot` allows visualizations including:

- Text frequencies
- Text correlations
- Text cooccurrences
- Text clusters
- Dependency parsing results

# Example visualisations

## Dependency Parser

### Example 1

```{r eval=FALSE}
model <- udpipe_download_model(language = "english", model_dir = "data/")
udmodel_english <- udpipe_load_model(file = "data/english-ewt-ud-2.4-190531.udpipe")
```
```{r}
x <- udpipe("His speech about marshmallows in New York is utter bullshit", "english")
plt <- textplot_dependencyparser(x, size = 4)
plt
```

### Example 2

```{r}
x <- udpipe("UDPipe provides tokenization, tagging, lemmatization and dependency parsing of raw text", "english")
plt <- textplot_dependencyparser(x, size = 4)
plt
```

## Biterm Topic Model plots

### Example 1

```{r}
data(example_btm, package = 'textplot')
model <- example_btm
plt <- plot(model, title = "BTM model", top_n = 5)
plt
```

```{r}
plt <- plot(model, title = "Biterm topic model", subtitle = "Topics 2 to 8", which = 2:8, top_n = 7)
plt
```

### Example 2

```{r message=FALSE, warning=FALSE}
library(data.table)

## Annotate text with parts of speech tags
data("brussels_reviews", package = "udpipe")
anno <- subset(brussels_reviews, language %in% "nl")
anno <- data.frame(doc_id = anno$id, text = anno$feedback, stringsAsFactors = FALSE)
anno <- udpipe(anno, "dutch")
## Get cooccurrences of nouns / adjectives and proper nouns
biterms <- as.data.table(anno)
biterms <- biterms[, cooccurrence(x = lemma, relevant = upos %in% c("NOUN", "PROPN", "ADJ"), skipgram = 2), by = list(doc_id)]
## Build the BTM model
set.seed(123456)
x <- subset(anno, upos %in% c("NOUN", "PROPN", "ADJ"))
x <- x[, c("doc_id", "lemma")]
model <- BTM(x, k = 5, beta = 0.01, iter = 2000, background = TRUE,
biterms = biterms, trace = 100)
plt <- plot(model)
plt
```

## Bar plots

Example showing frequency of adjectives.

```{r}
data("brussels_reviews_anno", package = "udpipe")
x <- subset(brussels_reviews_anno, xpos %in% "JJ")
x <- sort(table(x$lemma))
plt <- textplot_bar(x, top = 20, panel = "Adjectives", xlab = "Frequency", col.panel = "lightblue", 
                    cextext = 0.75, addpct = TRUE, cexpct = 0.5)
plt
```

## Correlation of texts

Top correlations above a certain threshold

```{r}
dtm <- subset(anno, upos %in% "ADJ")
dtm <- document_term_frequencies(dtm, document = "doc_id", term = "lemma")
dtm <- document_term_matrix(dtm)
dtm <- dtm_remove_lowfreq(dtm, minfreq = 5)
textplot_correlation_lines(dtm, top_n = 25, threshold = 0.01, lwd = 5, label = TRUE)
```

Correlations which are non-zero after fitting a glasso model.

```{r}
dtm <- subset(anno, upos %in% "NOUN")
dtm <- document_term_frequencies(dtm, document = "doc_id", term = "token")
dtm <- document_term_matrix(dtm)
dtm <- dtm_remove_lowfreq(dtm, minfreq = 20)
dtm <- dtm_remove_tfidf(dtm, top = 100)
term_correlations <- dtm_cor(dtm)
textplot_correlation_glasso(term_correlations, exclude_zero = TRUE)
```

## Co-occurrence of texts

Example showing adjectives occurring in the same document.

```{r}
data(brussels_reviews_anno, package = 'udpipe')
x <- subset(brussels_reviews_anno, xpos %in% "JJ" & language %in% "fr")
x <- cooccurrence(x, group = "doc_id", term = "lemma")
plt <- textplot_cooccurrence(x, title = "Adjective co-occurrences", top_n = 25)
plt
```

