---
title: "udpipe Dependency Parsing"
output:
  prettydoc::html_pretty:
    theme: Architect
    highlight: github
---

```{r message=FALSE, warning=FALSE}
setwd("~/R/Complete")
if(!require(easypackages)){install.packages("easypackages")}
library(easypackages)
packages("udpipe","igraph", "ggraph", "ggplot", "dplyr", "kableExtra", prompt = FALSE)
```

> There is a good function for plotting that should be recalled to make the plotting fast and simple.

# Introduction

Dependency parsing is an NLP technique which provides to each word in a sentence the link to another word in the sentence, which is called it's syntactical head. This link between each 2 words furthermore has a certain type of relationship giving you further details about it.

`udpipe` provides such a dependency parser. With the output of dependency parsing, you can answer questions like

- What is the nominal subject of a text
- What is the object of a verb
- Which word modifies a noun
- What is the linked to negative words
- Which words are compound statements
- What are noun phrases, verb phrases in the text

Example

> His speech about marshmallows in New York is utter bullshit

Running this code, will provide you the dependency relationships among the words of the sentence in the columns `token_id`, `head_token_id` and `dep_rel.` The possible values in the field `dep_rel` are defined at https://universaldependencies.org/u/dep/index.html.

```{r message=FALSE}
x <- udpipe("His speech about marshmallows in New York is utter bullshit", "english")

kable(select(x, -c(doc_id, paragraph_id, sentence_id, sentence))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

For visualizing the relationships between the words which were found, use 'ggraph'. 

```{r}
# library(igraph)
# library(ggraph)
# library(ggplot2)

plot_annotation <- function(x, size = 3){
  stopifnot(is.data.frame(x) & all(c("sentence_id", "token_id", "head_token_id", "dep_rel",
                                     "token_id", "token", "lemma", "upos", "xpos", "feats") %in% colnames(x)))
  x <- x[!is.na(x$head_token_id), ]
  x <- x[x$sentence_id %in% min(x$sentence_id), ]
  edges <- x[x$head_token_id != 0, c("token_id", "head_token_id", "dep_rel")]
  edges$label <- edges$dep_rel
  g <- graph_from_data_frame(edges,
                             vertices = x[, c("token_id", "token", "lemma", "upos", "xpos", "feats")],
                             directed = TRUE)
  ggraph(g, layout = "linear") +
    geom_edge_arc(ggplot2::aes(label = dep_rel, vjust = -0.20),
                  arrow = grid::arrow(length = unit(4, 'mm'), ends = "last", type = "closed"),
                  end_cap = ggraph::label_rect("wordswordswords"),
                  label_colour = "red", check_overlap = TRUE, label_size = size) +
    geom_node_label(ggplot2::aes(label = token), col = "darkgreen", size = size, fontface = "bold") +
    geom_node_text(ggplot2::aes(label = upos), nudge_y = -0.35, size = size) +
    theme_graph(base_family = "Arial Narrow") +
    labs(title = "udpipe output", subtitle = "tokenisation, parts of speech tagging & dependency relations")
}
```

```{r warning=FALSE}
plot_annotation(x, size = 4)
```

```{r warning=FALSE}
x <- udpipe("The economy is weak but the outlook is bright", "english")
plot_annotation(x, size = 4)
```

With dependency parsing you can answer the question _What is weak?_, it is the economy. _What is bright?_, it is the outlook as these nouns relate to the adjectives with nominal subject as type of relationship. That is more information than just looking at wordclouds.

Reference:  http://www.bnosac.be/index.php/blog/93-dependency-parsing-with-udpipe