---
title: "Feature Engineering"
output: html_document
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(!require(easypackages)){install.packages("easypackages")}
library(easypackages)
packages("readr", "lubridate", "dplyr", "ggplot2", "infotheo", prompt = FALSE)
```

## Engineering Dates

Just like we pulled the first word, word count and character count out of text features, we can also do some basic engineering on numerical and date features.

### Dates

A date field can be cast to an integer representation. Day 0 is 1/1/1970 (beginning of Unix time) - as you can imagine, this is very useful for modeling.

```{r}
print(as.numeric(as.Date('1970-01-01')))
```

But a date can yield a lot more data than just its integer representation. We’ll use the lubridate library to assist our extractions:

- Extract day, month, and short and long year
- Day count in year
- Day of the week
- Weekend
- Quarter

> Note: you will notice that there is an optional parameter to remove the original date. Some visualization
and modeling tools can handle dates automatically but for simplicity here, we want our entire data set to
be numerical. If that is an issue or if you want to retain the date for visualization and/or reporting, simply
turn the remove_original_date off.

```{r}
Feature_Engineer_Dates <- function(data_set, remove_original_date=TRUE) {
     data_set <- data.frame(data_set)
     date_features <- names(data_set[sapply(data_set, is.Date)])#is.Date is from lubridate
     for (feature_name in date_features) {
          data_set[,paste0(feature_name,'_DateInt')] <- as.numeric(data_set[,feature_name])
          data_set[,paste0(feature_name,'_Month')] <- as.integer(format(data_set[, feature_name], "%m"))
          data_set[,paste0(feature_name,'_ShortYear')] <- as.integer(format(data_set[,feature_name], "%y"))
          data_set[,paste0(feature_name,'_LongYear')] <- as.integer(format(data_set[,feature_name], "%Y"))
          data_set[,paste0(feature_name,'_Day')] <- as.integer(format(data_set[,feature_name], "%d"))
          # week day number requires first pulling the weekday label, 
          # creating the 7 week day levels, and casting to integer
          data_set[,paste0(feature_name,'_WeekDayNumber')] <- as.factor(weekdays(data_set[, feature_name]))
          
          levels(data_set[, paste0(feature_name, '_WeekDayNumber')]) <- 
               list(Monday=1, Tuesday=2, Wednesday=3, Thursday=4, Friday=5, Saturday=6, Sunday=7)
          
          data_set[, paste0(feature_name,'_WeekDayNumber')] <- 
               as.integer(data_set[,paste0(feature_name,'_WeekDayNumber')])
          
          data_set[,paste0(feature_name,'_IsWeekend')] <-  
               as.numeric(grepl("Saturday|Sunday", weekdays(data_set[, feature_name])))
          
          data_set[,paste0(feature_name,'_YearDayCount')] <- yday(data_set[,feature_name])
          data_set[,paste0(feature_name,'_Quarter')] <- lubridate::quarter(data_set[, feature_name], with_year = FALSE)
          data_set[,paste0(feature_name,'_Quarter')] <- lubridate::quarter(data_set[, feature_name], with_year = TRUE)
          if (remove_original_date)#delete the original date column only if remove_original_date=TRUE in function call
               data_set[, feature_name] <- NULL
     }
     return(data_set)
}
```

### R Code Walkthrough

Develop the data to explore the walkthrough:

```{r}
data_set <- data.frame(id=c(10,20,30,40,50), gender=c('male','female','female','male','female'), 
                some_date=c('2012-01-12','2012-01-12','2012-12-01','2012-05-30','2013-12-12'), 
                value=c(12.34, 32.2, 24.3, 83.1, 8.32), outcome=c(1,1,0,0,0))
data_set$some_date <- as.Date(data_set$some_date)#this would happen automatically using the pipeline.  Otherwise cast it
feature_name <- "some_date"
```

Create new features:

```{r}
data_set[,paste0(feature_name,'_DateInt')] <- as.numeric(data_set[,feature_name])#Days since 1970 (Unix Time)
glimpse(data_set)
data_set[,paste0(feature_name,'_Month')] <- as.integer(format(data_set[, feature_name], "%m"))#month
data_set[,paste0(feature_name,'_ShortYear')] <- as.integer(format(data_set[,feature_name], "%y"))# 2 digit year
data_set[,paste0(feature_name,'_LongYear')] <- as.integer(format(data_set[,feature_name], "%Y"))# 4 digit year
data_set[,paste0(feature_name,'_Day')] <- as.integer(format(data_set[,feature_name], "%d"))# day of month (1-31)
glimpse(data_set)
```

Calculate weekday number.  A bit more complicated because of factors, levels and ultimately an integer.

```{r}
#Make date col a factor.  Keeping original col untouched
data_set[,paste0(feature_name,'_WeekDayNumberTXT')] <- as.factor(weekdays(data_set[, feature_name]))
data_set$some_date_WeekDayNumber
#Define labels for facto column created above
levels(data_set[, paste0(feature_name, '_WeekDayNumberTXT')]) <- 
     list(Monday=1, Tuesday=2, Wednesday=3, Thursday=4, Friday=5, Saturday=6, Sunday=7)
data_set$some_date_WeekDayNumber
#Replace weekday name with corresponding Level integer value
data_set[, paste0(feature_name,'_WeekDayNumberINT')] <- as.integer(data_set[,paste0(feature_name,'_WeekDayNumberTXT')])
glimpse(data_set)
```

Obviously we would not keep both the Weekday name and weekday integer values in a model!

It is often useful to identify weekend days:

```{r}
#recall grepl returns TRUE/FALSE
data_set[,paste0(feature_name,'_IsWeekend')] <- as.numeric(grepl("Saturday|Sunday", weekdays(data_set[, feature_name])))
data_set$some_date_IsWeekend
table(data_set$some_date_WeekDayNumberTXT, data_set$some_date_IsWeekend)
```


```{r}
data_set[,paste0(feature_name,'_YearDayCount')] <- yday(data_set[,feature_name])
data_set$some_date_YearDayCount#the number of days since Jan 1
data_set[,paste0(feature_name,'_Quarter')] <- lubridate::quarter(data_set[, feature_name], with_year = FALSE)#
data_set$some_date_Quarter
data_set[,paste0(feature_name,'_Quarter')] <- lubridate::quarter(data_set[, feature_name], with_year = TRUE)
data_set$some_date_Quarter
glimpse(data_set)
```

Again, choose between either a quarter as eith a quater integer or one prefixed with the year - not both!

Test the function:

```{r}
mix_dataset <- data.frame(id=c(10,20,30,40,50), gender=c('male','female','female','male','female'), 
                some_date=c('2012-01-12','2012-01-12','2012-12-01','2012-05-30','2013-12-12'), 
                value=c(12.34, 32.2, 24.3, 83.1, 8.32), outcome=c(1,1,0,0,0))

write_csv(mix_dataset, './data/mix_dataset.csv')
mix_dataset <- read_csv('./data/mix_dataset.csv')
mix_dataset <- Feature_Engineer_Dates(mix_dataset)
head(mix_dataset)
```

> If you replaced read_csv with either read.csv or data.table fread, the some_date variable would not be cast as a Date!!

```{r}
mix_dataset_tmp <- read.csv('./data/mix_dataset.csv')
str(mix_dataset_tmp)#date as factor

mix_dataset_tmp <- data.table::fread('./data/mix_dataset.csv')
str(mix_dataset_tmp)#date a character
```

## Numerical Engineering

### Integers

First thing we have to do to extract additional intelligence out of an integer is to verify that it actually is an integer:

```{r}
print(is.integer(1))
print(class(1))
print(class(1L))
```

We cannot count on the is.integer function as it requires the value to be declared as an integer literal (L). Instead we will use the round function (in R 3.3.3 and up you can use `is.wholenumber`). We’ll explore some simple feature engineering to capture:

- Is feature equal to zero
- Is feature positive
- Binning feature values

```{r}
mix_dataset <- data.frame(id=c(1,2,3,4,5), mood=c(0,20,20,40,50), value=c(12.34, 32.2, 24.3, 83.1, 8.32), 
                          outcome=c(1,1,0,0,0))

write_csv(mix_dataset, './data/mix_dataset.csv')
mix_dataset <- read_csv('./data/mix_dataset.csv')
```

```{r}
Feature_Engineer_Integers <- function(data_set, features_to_ignore=c()) {
     data_set <- data.frame(data_set)
     for (feature_name in setdiff(names(data_set), features_to_ignore)) {
          if (class(data_set[,feature_name])=='numeric' | class(data_set[,feature_name])=='integer') {
               feature_vector <- data_set[,feature_name]
               if (all((feature_vector - round(feature_vector)) == 0)) {
                    # make sure we have more than 2 values excluding NAs
                    if (nrow(data_set %>% filter_(!is.na(feature_name)) %>% distinct_(feature_name)) > 2) {
                         #print(feature_name)#not really needed/helpful
                         data_set[,paste0(feature_name,'_IsZero')] <- ifelse(data_set[,feature_name]==0,1,0)
                         data_set[,paste0(feature_name,'_IsPositive')] <- ifelse(data_set[,feature_name]>=0,1,0)
                         # separate data into two bins using infotheo
                         data_discretized <- discretize(data_set[,feature_name], disc='equalfreq', nbins=2)
                         data_set[,paste0(feature_name,'_2Bins')] <- data_discretized$X
                         if (nrow(data_set %>% filter_(!is.na(feature_name)) %>% distinct_(feature_name)) > 4) {
                              # try 4 bins
                              data_discretized <- discretize(data_set[,feature_name], disc='equalfreq', nbins=4)
                              data_set[,paste0(feature_name,'_4Bins')] <- data_discretized$X
                         }
                    }
               }
          }
     }
     return (data_set)
}
```

> Note the nonstandard evaluation used in the dplyr statements above.  https://stackoverflow.com/questions/24569154/use-variable-names-in-functions-of-dplyr

### R Code Notes

- `if (all((feature_vector - round(feature_vector)) == 0))` simply ensures all the values in the column are integers - simple and fast!
- the `infotheo package` easily creates bins of data.  I might prefer the `binr package`.

```{r}
data_set <- mix_dataset
feature_name <- "mood"

data_discretized <- discretize(data_set[, feature_name], disc='equalfreq', nbins=2)
data_discretized$mood
```

Now lets test it all at once:

```{r}
mix_dataset <- read_csv('./data/mix_dataset.csv')
Feature_Engineer_Integers(mix_dataset, features_to_ignore=c('id'))
```

### Numbers

Feature engineering of numbers is an enormous subject that we’ll keep under control here. Most feature engineering should come out of the business context, something we can automate here. Here, we’ll look at some simple transformations that are applicable to a lot of data sets on whole/real numbers:

```{r}
Feature_Engineer_Numbers <- function(data_set, features_to_ignore=c()) {
     data_set <- data.frame(data_set)
     date_features <- setdiff(names(data_set[sapply(data_set, is.numeric)]), features_to_ignore)
     for (feature_name in date_features) {
          feature_vector <- data_set[,feature_name]
          if (is.integer(feature_vector) | is.numeric(feature_vector)) {
               if (any((feature_vector - round(feature_vector)) != 0)) {
                    # make sure we have more than 2 values excluding NAs
                    if (nrow(data_set %>% filter_(!is.na(feature_name)) %>% distinct_(feature_name)) > 2) {
                         # print(feature_name)#Not really helpful
                         # polynomial transformation
                         poly_vector <- poly(x=feature_vector, degree = 2)
                         data_set[,paste0(feature_name, "_poly1")] <- poly_vector[,1]
                         data_set[,paste0(feature_name, "_poly2")] <- poly_vector[,2]
                         # log transform
                         data_set[,paste0(feature_name, "_log")] <- log(x = feature_vector)
                         # exponential transform
                         data_set[,paste0(feature_name, "_exp")] <- exp(x = feature_vector)
                         # rounding
                         data_set[,paste0(feature_name, "_rnd")] <- round(x = feature_vector, digits = 0)
                         # binning into 2 bins
                         data_discretized <- discretize(data_set[,feature_name], disc='equalfreq', nbins=2)
                         data_set[,paste0(feature_name,'_2Bins')] <- data_discretized$X
                    }
               }
          }
     }
     return(data_set)
}
```


```{r}

mix_dataset <- data.frame(id=sample(1:100, 100, replace=F), value=runif(100, 1.0, 55.5))
write_csv(mix_dataset, './data/mix_dataset.csv')
mix_dataset <- read_csv('./data/mix_dataset.csv')
head(Feature_Engineer_Numbers(mix_dataset, features_to_ignore=c()))
```

The Feature_Engineer_Numbers function will only transform features containing real numbers. It then applies a 2-degree polynomial transform, a simple log and exponential transform. It also rounds the data and splits it into two buckets using library infotheo. All these transformations are highly customizable, you could try 3-degree polynomial transform, round only to the 1st or 2nd digit. You could split the data into many more bins. Depending on the data, a few things can break, whether or not you have negative numbers, and too much or too little variation. This would definitely be an ideal candidate for a try/ catch error handling.

For more information on the following transformer functions:

- Log, Exp (http://www.inside-r.org/r-doc/base/log)
- poly/log (https://stat.ethz.ch/R-manual/R-devel/library/stats/html/poly.html)
- discretize (http://www.inside-r.org/packages/cran/infotheo/docs/discretize)

## Pipeline Check

```{r}
remove(list=ls())
source("../EDA/PipelineFunctions.R")

mix_dataset <- data.frame(id=c(1,2,3,4,5), gender=c('male','female','female','male','female'), 
                some_date=c('2012-01-01','2013-01-01','2014-01-01','2015-01-01','20 16-01-01'), 
                mood=c(0,20,20,NA,50), value=c(12.34, 32.2, 24.3, 83.1, 8.32), 
                outcome=c(1,1,0,0,0)) 
 
write_csv(mix_dataset, '../EDA/data/mix_dataset.csv') 
mix_dataset <- as.data.frame(read_csv('../EDA/data/mix_dataset.csv')) 
 
# automated pipeline 
mix_dataset <- Get_Free_Text_Measures(data_set = mix_dataset)#no change since there is no free text
mix_dataset <- Binarize_Features(data_set = mix_dataset, leave_out_one_level = TRUE)#changes Gender
mix_dataset <- Impute_Features(data_set = mix_dataset, use_mean_instead_of_0 = TRUE)#Replaces the NA
mix_dataset <- Feature_Engineer_Dates(data_set = mix_dataset)#Extract detail from Data variable
mix_dataset <- Feature_Engineer_Integers(data_set = mix_dataset)#Auto Bin the integers-many cols added
mix_dataset <- Feature_Engineer_Numbers(data_set = mix_dataset)#creates some Inf because not enough numbers to run the exp in this case.  Need to run Impute again to clean up the new colmns afeter feature engineering

head(mix_dataset,2)
summary(mix_dataset)
```
