---
title: "EDA - Reading Data"
output: html_document
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(!require(easypackages)){install.packages("easypackages")}
library(easypackages)
packages("readr", "data.table", "dplyr", "ggplot2", prompt = FALSE)
```

## Reading Data

### readLines

A first investigative exploration of a data set is the readLines function (https://stat.ethz.ch/R-manual/Rdevel/library/base/html/readLines.html). It allows you to cull a small amount of lines from the top of a file, no matter how big the file is. This is a silly example but imagine if your file is over 10 gigabytes in size, why waste time and memory when you aren’t sure what it contains?

Let’s use the readLines function to open part of a text file off CRAN:

```{r}
readLines('https://cran.r-project.org/src/base/README', n=20)
```

This is a great way of discovering what data types are contained in a very large data set. You will see in the
next few functions we use, knowing the data type in advance can speed things up tremendously.

### read.table, read.csv, read.csv2

read.csv is the most common reader in R, followed closely by read.table. In fact, read.csv and read.csv2 are both wrappers over read.table (https://stat.ethz.ch/R-manual/Rdevel/library/utils/html/read.table.html).

```{r eval=FALSE}
read.table(file, header = FALSE, sep = "", quote = "\"'",
     dec = ".", numerals = c("allow.loss", "warn.loss", "no.loss"),
     row.names, col.names, as.is = !stringsAsFactors,
     na.strings = "NA", colClasses = NA, nrows = -1,
     skip = 0, check.names = TRUE, fill = !blank.lines.skip,
     strip.white = FALSE, blank.lines.skip = TRUE,
     comment.char = "#",
     allowEscapes = FALSE, flush = FALSE,
     stringsAsFactors = default.stringsAsFactors(),
     fileEncoding = "", encoding = "unknown", text, skipNul = FALSE)
```

Let’s use read.table. Here we’ll read the Titanic dataset from the University of Colorado Denver.  Eventhough it off the Internet, the same use applies to your hard-drive. We set the separator parameter of \\t for tabular and request to consider the first row as headers:

```{r}
Titanic_dataset <- read.table('http://math.ucdenver.edu/RTutorial/titanic.txt', sep='\t', header=TRUE)
head(Titanic_dataset)
```

Let’s see if the function figured out the data types contained in the Titanic dataset:

```{r}
str(Titanic_dataset)
```

Not bad, but feature Names shouldn’t be a factor but a character. In this case, we could change it after the case or force the reader to make all text a character instead of factor by setting the stringsAsFactors to FALSE:

```{r}
Titanic_dataset <- read.table('http://math.ucdenver.edu/RTutorial/titanic.txt', sep='\t', header=TRUE, stringsAsFactors=FALSE)
str(Titanic_dataset)
```

You can also pass custom column names directly to the read.table function:

```{r eval=FALSE}

actg320_colnames <- c('id','time','censor','time_d','censor_d','treatment','treatment_group','strat2','sex','raceth','ivdrug', 
                      'hemophil','karnof','cd4','priorzdv','age')
#Note - the file below is no longer shared
actg320 <- read.table('https://www.umass.edu/statdata/statdata/data/actg320.dat', col.names = actg320_colnames)
head(actg320)
```

Here’s a quick look at read.csv. For more differences between these readers see: read.table
(http://www.inside-r.org/r-doc/utils/read.table)

```{r}
Iris_dataset <- read.csv('http://archive.ics.uci.edu/ml/machine-learning-databases/iris/bezdekIris.data', header=FALSE)
head(Iris_dataset)
```

Before we move to more sophisticated readers, let’s build a simple data frame to work with. We’ll create a
very small data set made of the following data types: integers, factors, doubles, and dates.

```{r}
mix_dataset <- data.frame(id=c(10,20,30,40,50), gender=c('male','female','female','male','female'), 
                          some_date=c('2012-01-12','2012-01-12','2012-12-01','2012-05-30','2013-12-12'), 
                          value=c(12.34, 32.2, 24.3, 83.1, 8.32), outcome=c(1,1,0,0,0))
write.csv(mix_dataset, '../data/mix_dataset.csv', row.names = FALSE)
```

We save it to your current working directory and read it back again. Let’s look at it using read.csv:

```{r}
mix_dataset <- read.csv('../data/mix_dataset.csv', stringsAsFactors = FALSE)
str(mix_dataset)
```

##Heavy-duty Readers

### readr

Let’s look at some readers that aren’t part of the base package. readr {readr} (https://cran.rproject.org/web/packages/readr/README.html) is a relatively new package maintained by Hadley Wickham. It does a great job at inferring data types and is fast!

As you can see, some_date is correctly cast as Date, and it does pick up two integer fields:

```{r readr}
mix_dataset <- read_csv('./data/mix_dataset.csv')
str(mix_dataset)
```

If you know in advance the data types in a data set, you can pass it along to the function to save it time and processing. In the first read, we force some_date to character, in the read we force the id field to be a numeric instead of an integer.

```{r}
mix_dataset <- read_csv('./data/mix_dataset.csv', col_types='nccnn')#number, character, character, number, number
str(mix_dataset)

mix_dataset <- read_csv('./data/mix_dataset.csv', col_types='ncDni')#D is date
str(mix_dataset)
```

readr returns classes 'tbl_df', 'tbl', 'data.frame'.  From past experience, I have used readr only to find later in my code these multiple classes introduced code issues.  If that happens, you can still use eadr but cast it as a dataframe:

```{r}
mix_dataset <- read_csv('./data/mix_dataset.csv')
class(mix_dataset)

mix_dataset <- data.frame(read_csv('./data/mix_dataset.csv'))
class(mix_dataset)
```

### fread

fread {data.table} (http://www.inside-r.org/packages/cran/data.table/docs/fread) is the *fastest* of the bunch presented so far. **Set parameter data.table=FALSE to return a data frame**:

```{r datatable}
mix_dataset <- fread('./data/mix_dataset.csv', showProgress=TRUE, data.table=FALSE)
str(mix_dataset)
```

If you only want a subset of columns, you can choose them directly in the fread command using the select parameter (similarly you can use the drop parameter to remove features):

```{r datatable2}
mix_dataset <- fread('./data/mix_dataset.csv', data.table=FALSE, select = c('value', 'outcome'))
head(mix_dataset)
```

For reference, if you want to load data from an Excel spreadsheet, here are some popular libraries:

- XLConnect (https://cran.r-project.org/web/packages/XLConnect/index.html)
- openxlsx (https://cran.rproject.org/web/packages/openxlsx/index.html)
readxl (https://cran.rproject.org/web/packages/readxl/index.html).

##Pipeline Check

So far we don’t really need to create any custom code, simply use read_csv or fread if you have large data and/or complex data types, otherwise stick to read.csv. If this is new data, use the readLines function:

`path_and_file_name <- 'https://cran.r-project.org/src/base/README'`
`print(readLines(path_and_file_name, n=5))`

