---
title: "Selecting the Best Clustering Algorithm"
output: html_document
---

### Introduction

We will use the *clValid* package which can be used to simultaneously compare multiple clustering algorithms in a single function call to find the best algo and the optimal number of clusters.

The clValid package leverages two cluster validation methods:

1. Internal Measures: This includes connectivity, the silhouette coefficient and the Dunn Index.
2. Stability Measures:  This is a special version on internal measures evaluating the consistency of a clustering result by comparing it with the clusters obtained after each column is removed one at a time.  This includes detail which I will not detail:

     - APN - Average Proportion of Non-Overlap
     - AD - Average Distance
     - ADM - Average Distance Between Means
     - FOM - Figure of Merit
     
### clValid Selection Example

```{r getPackage, echo=FALSE, warning=FALSE, message=FALSE}
if(!require(easypackages)){install.packages("easypackages")}
library(easypackages)
packages("clValid",  prompt = FALSE)
```

```{r try_clValid, message=FALSE, warning=FALSE}
df <- scale(iris[, -5])

#Compute clValid
clmethods <- c("hierarchical", "kmeans", "pam")
internal <- clValid(df, nClust=2:6, clMethods=clmethods, validation="internal")
summary(internal)
```

The summary above suggests that hierarchical clustering with 2 clusters performs the best in each case.  Regardless of the clustering algorithm, the optimal number of clusters appears to be 2 using these 3 measures.  (Of course we know that there are really 3 clusters).

Stability measure can be used like this:

```{r}
#clmethods <- c("hierarchical", "kmeans", "pam")
stab <- clValid(df, nClust=2:6, clmethods=clmethods, validation="stability")
optimalScores(stab)
```

The results above seem to confuse the number of clusters even more.  APN and ADM suggest 2 clusters and the others 6.  Hmm, not sure what to make of this package!
