---
title: 'Using the BreakDown Package'
output:
    rmdformats::readthedown:
      highlight: pygments
      code_folding: show
---
<style type="text/css">
p{ /* Normal  */
   font-size: 14px;
   line-height: 18px;
}
body{ /* Normal  */
   font-size: 14px;
}
td {  /* Table  */
   font-size: 12px;
}
h1 { /* Header 1 */
font-size: 26px;
color: #4294ce;
}
h2 { /* Header 2 */
font-size: 22px;
}
h3 { /* Header 3 */
font-size: 18px;
}
code.r{ /* Code block */
  font-size: 12px;
}
pre { /* Code block */
  font-size: 12px
}
#table-of-contents h2 {
background-color: #4294ce;
}
#table-of-contents{
background: #688FAD;
}
#nav-top span.glyphicon{
color: #4294ce;
}
#postamble{
background: #4294ce;
border-top: ;
}
</style>

# Introduction

`breakDown` prcreates model agnostic waterfall charts to make model results more interprettable.

The idea behind Break Down Plots it to decompose model prediction for a *single observation*. 

Plots will work for binary classifiers and general regression models. 

# Load Libraries

```{r loadLibs1, warning=FALSE, message=FALSE}
#if(!require(breakDown)){devtools::install_github("pbiecek/breakDown")}
if(!require(easypackages)){install.packages("easypackages")}
library(easypackages)
packages("plyr","dplyr","ggplot2", "readr", "tidyr", "gridExtra", "stringr", "lubridate", 
        "caret", "gbm", "modelr", "ggthemes", "breakDown", prompt = FALSE)

options(scipen = 999)#Do not display exponents
```

# Get Data

`HR_Data` is now included in the `breakDown` package.

```{r}
head(HR_data, 3)
```

# breakDown Examples

## randomForest

```{r message=FALSE}
library("randomForest")
model_rf <- randomForest(factor(left)~., data = HR_data, family = "binomial", maxnodes = 5)

predict.function <- function(model_rf, new_observation) 
predict(model_rf, new_observation, type="prob")[,2]
predict.function(model_rf, HR_data[11,-7])
```
```{r}
explain_rf <- broken(model_rf, HR_data[11,-7], data = HR_data[,-7], predict.function = predict.function, 
                    direction = "down")
explain_rf
```
```{r}
plot(explain_rf) + ggtitle("breakDown plot  (direction=down) for randomForest model")
```

```{r}
explain_2 <- broken(model_rf, HR_data[11,-7], data = HR_data[,-7],
                    predict.function = predict.function, direction = "up")

plot(explain_2) + ggtitle("breakDown plot (direction=up) for randomForest model")
```

## ranger Example

Now let’s create a ranger classification forest for churn, the left variable.

```{r message=FALSE}
library(ranger)
HR_data$left <- factor(HR_data$left)
model_ranger <- ranger(left ~ ., data = HR_data, importance = 'impurity', probability=TRUE, min.node.size = 2000)

predict.function <- function(model_ranger, new_observation) 
predict(model_ranger, new_observation, type = "response")$predictions[,2]

predict.function(model_ranger, HR_data[11,])
```
```{r}
explain_ranger <- broken(model_ranger, HR_data[11,-7], data = HR_data[,-7],
                    predict.function = predict.function, direction = "down")
explain_ranger
```
```{r}
plot(explain_ranger) + ggtitle("breakDown plot  (direction=down) for ranger model")
```

## GLM Logistic Example

```{r}
model_glm <- glm(left~., data = HR_data, family = "binomial")

predict(model_glm, HR_data[11,], type = "link")
```
```{r}
explain_glm <- broken(model_glm, HR_data[11,])
explain_glm
```
```{r}
plot(explain_glm) + ggtitle("breakDown plot for linear predictors")
```

## Linear Regression

```{r}
wine <- read.table("./data/winequality-white.csv", header = T, sep=";")
head(wine, 3)
```
```{r}
model_lm <- lm(quality ~ fixed.acidity + volatile.acidity + citric.acid + residual.sugar + 
            chlorides + free.sulfur.dioxide + total.sulfur.dioxide + density + pH + sulphates + alcohol,
            data = wine)
new_observation <- wine[1,]
br <- broken(model_lm, new_observation)
br
```
```{r}
plot(br)
```

## SVM Example

SVM using `kernlab` and `e1071` packages.

```{r message=FALSE}
library(kernlab)

wine_svm_model_kern <- ksvm(quality~., data = wine)
wine_svm_model_kern
```
```{r}
library(e1071)

wine_svm_model_e <- svm(quality~., data = wine)
wine_svm_model_e
```

Now we are ready to call the `broken()` function. Since `kernlab` is using S4 methods we need to pass the hook to `kernlab:::predict` method.

```{r}
nobs <- wine[5, , drop = FALSE]
base_prediction1 <- predict(wine_svm_model_kern, nobs)
set.seed(1313)

explain_5_kern <- broken(wine_svm_model_kern, new_observation = nobs, 
                    data = wine, predict.function = predict,
                    baseline = "intercept", direction = "up")
explain_5_kern
```
```{r}
plot(explain_5_kern) + ggtitle(paste0("Prediction for SVM kernLab model ", round(base_prediction1, 3)))
```

```{r}
base_prediction2 <- predict(wine_svm_model_e, nobs)
set.seed(1313)

explain_5_e <- broken(wine_svm_model_e, new_observation = nobs, 
                    data = wine, predict.function = predict,
                    baseline = "intercept", direction = "up")
explain_5_e
```
```{r}
plot(explain_5_e) + ggtitle(paste0("Prediction for SVM e1071 model ", round(base_prediction2, 3)))
```

## xgBoost Example

```{r message=FALSE}
library(xgboost)

model_matrix_train <- model.matrix(left ~ . - 1, HR_data)
data_train <- xgb.DMatrix(model_matrix_train, label = as.numeric(HR_data$left))
param <- list(objective = "reg:linear")

HR_xgb_model <- xgb.train(param, data_train, nrounds = 50)
HR_xgb_model
```
```{r eval=FALSE}
nobs <- model_matrix_train[1L, , drop = FALSE]

explain_xgb <- broken(HR_xgb_model, new_observation = nobs, data = model_matrix_train)#require breakDown 0.15 or higher
explain_xgb
```

```{r eval=FALSE}
plot(explain_xgb) + ggtitle("breakDown plot for xgboost model")
```


##caret Model Example

This uses a GLM inside `caret`.

```{r}
library(caret)

set.seed(2)
training <- twoClassSim(50, linearVars = 2)
#twoClassSim simulates regression and classification data with truly important predictors and irrelevant predictions
trainX <- training[, -ncol(training)]
trainY <- training$Class

head(training)
```

```{r}
cctrl1 <- trainControl(method = "cv", number = 3, returnResamp = "all",
                       classProbs = TRUE, summaryFunction = twoClassSummary)

test_class_cv_model <- train(trainX, trainY, method = "glm", trControl = cctrl1,
                             metric = "ROC", preProc = c("center", "scale"))
test_class_cv_model
```

To use `breakDown` need a function that calculates scores/predictions for a single observation. By default the `predict()` returns predicted class.

Add `type = "prob"` argument to get scores. And since there will be two scores for each observarion we need to extract one of them.

```{r}
predict.fun <- function(model, x) predict(model, x, type = "prob")[,1]
testing <- twoClassSim(10, linearVars = 2)
predict.fun(test_class_cv_model, testing[1,])
```

```{r}
library("breakDown")
explain_caret <- broken(test_class_cv_model, testing[1,], data = trainX, predict.function = predict.fun)
explain_caret
```

```{r}
plot(explain_caret) + ggtitle("breakDown plot for caret/glm model")
```


